---
title: "Project 1 Time Series"
output: html_notebook
author: "Halan Badilla Osorio"
---


Dataset & Data Preprocessing
    Source and describe the dataset(s) used.
    Conduct exploratory data analysis (EDA), including visualization.
    Check for stationarity (Augmented Dickey-Fuller test, ACF/PACF analysis).
Modeling & Analysis
    Fit a distributed lag model and interpret results.
    Optionally, consider alternative models (VAR, ARDL, or intervention analysis).
    Evaluate model performance and discuss assumptions.
Written Report (5-10 pages, including visualizations)
    Introduction: Research question, why it matters.
    Methodology: Data collection, preprocessing, modeling approach.
    Results: Key findings with interpretation.
    Conclusion: Policy/business implications, limitations, future research.

The effect of government stimulus on employment and GDP
Question: Do government stimulus progams lead to higher employment and GDP growth?
Data Sources; FRED
  GDP Growth Rate
  Individual Income Tax: Tax Rates for Regular Tax: Highest Bracket
  Labor Force Participation Rate
  Labor Force Participation Rate: Women 
  Labor Force Participation Rate: Men
  Median Household Income
  
What to do: Distributed Lag Model to assess whether tax cuts impact economic growth and income inequality over time.

```{r}
library(dplyr)
library(tidyverse)
library(readxl)
library(tsibble)
library(fable)
```


```{r}
setwd("C:/Users/halan/Desktop/FPU2025Spring/Time Series/Project 1")
GDP <- read_csv("GDP.csv")
GDPGrowthRate <- read_csv("GDPGrowthRate.csv")
IIT <- read_csv("IITTRHB.csv")
LForce <- read_csv("LaborForceParticipationRate.csv")
LForceMen <- read_csv("LaborForceParticipationRateMen.csv")
LForceWomen <- read_csv("LaborForceParticipationRateWomen.csv")
MedianHousehold <- read_csv("RealMedianHouseholdIncome.csv")
```
```{r}
# Renaming the columns
LForce <- rename(LForce, Date = observation_date, Rate = CIVPART)
LForceMen <- rename(LForceMen, Date = observation_date, Rate = LNS11300001)
LForceWomen <- rename(LForceWomen, Date = observation_date, Rate = LNS11300002)
```

```{r}
# Giving a color to each to then merge
LForce <- LForce %>%
  mutate(Type = "Both")

LForceMen <- LForceMen %>%
  mutate(Type = "Men")

LForceWomen <- LForceWomen %>%
  mutate(Type = "Women")

# Full joining them
LForceMerged <- rbind(LForce, LForceMen, LForceWomen)

# Tsibble
LForceMerged <- LForceMerged %>%
  as_tsibble(index = Date, key = Type)
```

```{r}
autoplot(LForceMerged, Rate) +
  labs(title = "Labor Force %", x = "Time (Years & Month)", y = "Rate (%)") +
  scale_color_manual(values = c("Purple", "Blue", "Red"))
```

